\documentclass[tikz,border=10pt]{article}
\newcommand{\course}{Physics}
\usepackage[margin=1.5in]{geometry}
\newcommand{\instructor}{Athan Zhang \& Jeffrey Chen}
\usepackage{amsmath}
\usepackage{pgfplots}
\pgfplotsset{width=10cm, compat=1.9}
\usepackage[inline]{enumitem}
\pgfplotsset{compat=1.12}
\usepackage{float}
\usepackage{parskip}

% Document Body
\begin{document}

% Title
\begin{center}
\textbf{\Huge Problem Set 02} % Replace "#" with the problem set number
\end{center}

% Course and Instructor Information
\begin{flushleft}
\emph{Course:} \course \\
\emph{Instructors:} \instructor \\
\end{flushleft}

\section*{Problem 1}

Three Vectors $\Vec{A}, \Vec{B}, \Vec{C}$ have the following components

\begin{table}[H]
    \centering
    \begin{tabular}{c|c c c}
          & $\Vec{A}$ & $\Vec{B}$ & $\Vec{C}$ \\
          \hline
        x & +6 & -3 & -2 \\
        y & +3 & 0 & +2 \\
    \end{tabular}
\end{table}

Find the following:

\begin{enumerate}
    \item Magnitude of $\Vec{A} - \Vec{B}$
    \item Magnitude of $\Vec{A} + \Vec{B} + \Vec{C}$
    \item The Angle $\Vec{A}$ makes with the x-axis
    \item The Angle $\Vec{A} + \Vec{B} + \Vec{C}$ makes with the x-axis
\end{enumerate}

\section*{Problem 2}

The faces of a cube of side 3m are parallel to the coordinate planes with one corner at the origin. An ant begins at the origin and walks along the three edges until it is at the far corner. Write the displacement vector of the fly using the unit vectors $\hat{i}, \hat{j}, \hat{k}$ and find the magnitude of this displacement. What was the distance traveled by the ant?

\section*{Problem 3}
A river is 0.76 km wide. The banks are straight and parallel. The current is 5.0 km/h and is parallel to the banks. A boat has a maximum speed of 3 km/h in still water. The pilot of the boat wishes to go in a straight line across the banks such that his line of travel is perpendicular to the banks. The pilot should head at what angle upstream?

\section*{Problem 4}
A new swimmer bravely heads directly across a river, swimming at 1.6 m/s relative to still water. She arrives at a point 40 m downstream from the point directly across the river, which is 80m wide. 
\begin{enumerate}
    \item What is the speed of the river current?
    \item What is the swimmer's speed relative to the shore?
    \item In what direction should the swimmer head so as to arrive at the point directly opposite her starting point?
\end{enumerate}

\section*{Problem 5}
A car is traveling due north with a velocity of 50 km/h. At the same time, a truck is traveling due west with a velocity of 40 km/h. Determine the resultant velocity of the car relative to the truck. Draw a diagram with the car and trunk originating from the same point. Express your answer in vector component form.

\section*{Problem 6}
Two cyclists start from the same point and travel in opposite directions. Cyclist A rides at a speed of 20 km/h, while Cyclist B rides at a rate of 15 km/h. If they are initially 30 km apart, how long will it take for them to be 120 km apart?

\section*{Problem 7}
It is the year 3023, and humans have developed near-lightspeed travel. X Æ A-Xii is a pilot of the Gaian Intergalactic fleet aboard the S.S. Big Bird. The S.S. Big Bird travels at a maximum speed of $2.71 \times 10^{7}$ m/s. In a battle, an enemy spaceship chases the badly damaged but fully functional S.S. Big Bird at a speed of $1.32 \times 10^{7}$ m/s relative to the S.S. Big Bird.
\begin{enumerate}
    \item What is the speed of the enemy spaceship relative to space?
    \item Calculate the difference between this speed and the speed from an approximate calculation.
\end{enumerate}

\section*{Problem 8}
Two boat landings are 2.0 km apart on the same bank of a stream that flows at 1.4 km/h parallel to the bank. A motorboat makes a round trip between the two landings in 50min. What is the speed of the boat relative to still water? 

\section*{Problem 9}
A pool ball leaves a 0.60-meter high table with an initial horizontal velocity of 2.4 m/s. Calculate the time required for the pool ball to fall to the ground and the horizontal distance between the table's edge and the ball's landing location.

\section*{Problem 10}
A football is kicked with an initial velocity of 25 m/s at an angle of 45 degrees with the horizontal. Determine the time of flight, the horizontal distance, and the peak height of the football.

\section*{Problem 11}
A long jumper leaves the ground with an initial velocity of 12 m/s at an angle of 28 degrees above the horizontal. Determine the time of flight, the horizontal distance, and the peak height of the long jumper.

\section*{Problem 12}
A projectile is fired with an initial speed of 53 m/s. Find the angle of projection such that the maximum height of the projectile is equal to its horizontal range.

\section*{Problem 13}
A ball thrown into the air lands 40m away 2.44 s later. Find the direction and magnitude of its initial velocity.

\section*{Problem 14}
A powerful rabbit jumps with an initial velocity of 7.2 m/s at an angle of 32 degrees from the horizontal from the very edge of a cliff that is 70 meters tall. 
\begin{enumerate}
    \item How long does the rabbit stay in the air?
    \item How far does the rabbit travel?
\end{enumerate}

\section*{Problem 15}
Lebron James throws a basketball at a vertical wall 5 meters away. Since Lebron is approximately 2 meters tall, the ball leaves his hand with an initial velocity $(10\hat{i} + 10\hat{j})$ m/s from 2 meters off the ground. When the ball hits the wall, the horizontal component of its velocity is flipped the vertical component remains unchanged. How far does the ball land from Lebron?

\end{document}
