\documentclass[tikz,border=10pt]{article}
\newcommand{\course}{Physics}
\usepackage[margin=1.5in]{geometry}
\newcommand{\instructor}{Athan Zhang \& Jeffrey Chen}
\usepackage{amsmath}
\usepackage{pgfplots}
\pgfplotsset{width=10cm, compat=1.9}
\usepackage[inline]{enumitem}
\pgfplotsset{compat=1.12}
\usepackage{float}
\usepackage{parskip}

% Document Body
\begin{document}

\begin{center}
\textbf{\Huge Problem Set 04} % Replace "#" with the problem set number
\end{center}

% Course and Instructor Information
\begin{flushleft}
\emph{Course:} \course \\
\emph{Instructors:} \instructor \\
\end{flushleft}

\section*{Problem 1}

A 20 N block rests on a horizontal surface. The coefficients of static and kinetic friction between the surface and the block are $\mu_s = 0.8$ and $\mu_k = 0.6$. A horizontal string is attached to the block and constant tension $T$ is maintained in the string. what is the force of friction acting on the block if
\begin{enumerate}
    \item $T = 15$ N
    \item $T = 20$ N
\end{enumerate}

\section*{Problem 2}
Jeffrey is an excellent and famous singer. During a performance, he makes a grand entrance by running out from behind the curtains and sliding on his knees across his the stage. His initial speed is 3 m/s. After sliding 2 m, he comes to rest . What is the coefficient of kinetic friction between Jeffrey and the stage?

\section*{Problem 3}
The coefficient of static friction between the bed of a truck and a box resting on it is 0.3. The truck is traveling at 80 km/h along a horizontal road. What is the least distance in which the truck can stop if the box is not to slide?

\section*{Problem 4}
A block of mass $m$ slides with initial speed $v_0$ on a horizontal surface. If the coefficient of kinetic friction between the block and the surface is $\mu_k$, find the distance $d$ that the block moves before coming to rest.

\section*{Problem 5}
A block of mass $m$ rests on a wedge of mass $M$, which makes an angle $\theta$ with a horizontal surface. A horizontal force $\Vec{F}$ acts on the wedge. There is no friction between the wedge and the horizontal surface. However, there is friction between the wedge and the block with a coefficient of static friction of $\mu_s = 0.8$. Find the maximum and minimum values of $F$ for which the block does not slip.

\section*{Problem 6}
A 0.2 kg stone is attached to a 0.8 m long string that is connected to a horizontal ceiling. The stone is rotating in a circle in the horizontal plane. The string makes an angle of 20 degrees with the flat ceiling. Determine the speed of the stone.

\section*{Problem 7}
A mass $m_1$ moves with speed $v$ in a circular path of radius $R$ on a frictionless table. It is attached to a string that passes through a hole at the center of the table. A second mass $m_2$ is attached to the other end of the string and is under the table, suspended in air. Derive an expression for $R$ in terms of $m_1$, $m_2$, and $v$.

\section*{Problem 8}
A 100 g disk sits on a horizontally rotating turntable. The turntable makes one revolution each second. The disk is located 10 cm from the axis of rotation on the turntable.
\begin{enumerate}
    \item What is the frictional force acting on the disk?
    \item The disk will slide off the turntable if it is located at a radius larger than 16 cm from the axis of rotation. What is the coefficient of static friction? (Hint: Remember you are given an angular velocity, that is, how many radians/rotations occur over a time, not a velocity)
\end{enumerate}

\section*{Problem 9}
On a banked curve of 15 degrees. A 2000 kg car drives along a circular road with a radius of 20 m. If the coefficient of static friction between the car's wheels and the road is $\mu_s = 0.3$, what is the maximum and minimum speed the car can go at before slipping (moving up and down the bank)?

\end{document}
