\documentclass[tikz,border=10pt]{article}
\newcommand{\course}{Calculus}
\usepackage[margin=1.5in]{geometry}
\newcommand{\instructor}{Athan Zhang \& Jeffrey Chen}
\usepackage{amsmath}
\usepackage{pgfplots}
\pgfplotsset{width=10cm, compat=1.9}
\usepackage[inline]{enumitem}

\usepgfplotslibrary{external}
\tikzexternalize
\pgfplotsset{compat=1.12}

% Document Body
\begin{document}

% Title
\begin{center}
\textbf{\Huge Problem Set 01} % Replace "#" with the problem set number
\end{center}

% Course and Instructor Information
\begin{flushleft}
\emph{Course:} \course \\
\emph{Instructors:} \instructor \\
\end{flushleft}

% Problem 1
\section*{Problem 1}
Does $\lim_{x \to 4} \frac{\sqrt{x}-2}{x-4}$ exist? If so, evaluate it. \textit{Hint:} try to factor the function.

\section*{Problem 2}
Categorize the end behavior of the function $\frac{e^x-e^{-x}}{2}$. 


\section*{Problem 3}
Find and categorize all the discontinuities of the function $\frac{5x^3}{x^3+2x^2-8x}$. 


\section*{Problem 4}
Find and categorize all the discontinuities of the function $cot(x)$. What about $sin(x)cot(x)$?


\section*{Problem 5}
Jake is driving his car. His velocity can be modeled as $t^3-t^2+10t$, where $t$ represents the number of seconds elapsed since he started driving. What is his average velocity over the first $10$ seconds of him driving?

\section*{Problem 6}
Sarah is driving her car. Her car's position can be modeled as $t^2+4t+10$, where $t$ represents the number of minutes elapsed since she started driving. By the time $t$ she has drove $42$ miles, what is her average velocity from the start to $t$?

\section*{Problem 7}
The population of Town A can be modeled as $P(t) = e^{0.3t}+500$, where $t$ represents the number of years since Town A's founding. What is the average rate of Town A's population growth over the first $10$ years since its founding, to the nearest hundredth?

\section*{Problem 8}
What is the slope of the line tangent to the curve $x^3+2x^2+1$ at $x=1$?

\section*{Problem 9}
Athan is driving at $2t^2+12t+2$ miles per hour, where $t$ represents the number of minutes he has been driving. By the time he reaches a velocity of $130$ miles per hour, what was his instantaneous acceleration exactly $3$ minutes ago?

\end{document}
